---
title: "Final Portfolio:"
author: "Tate McDonald"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

# BACKGROUND



# STUDY QUESTION & HYPOTHESIS

 
## Question  
Do infant vaccinations have a significant impact on the decline of global infant mortality rates over time?

## Hypothesis
Infants who receive available vaccinations at/near time of birth have a lower mortality rate than infants who do not receive such vaccinations.

## Predictions & Possible Visualizations
A scatterplot showing data from years 1974-2024 will have two variables: mortality rates of infants who receive vaccinations, and mortality rates of infants who do not receive vaccinations (deaths per 1,000 births). A linear regression model can be used to determine if there is a significant difference between mortality rates in infants who receive vaccines vs. those who do not. It is predicted that a linear regression model will have a small p-value showing that infants who receive vaccines are much less likely to contribute to the global infant mortality rates than infants who do not receive vaccines.


# METHODS

## Procedure


## First Analysis: Homework 10

```{r message=FALSE, warning=FALSE}
# Load necessary packages
library(tidyverse)

# Import data
data <- read_csv("Global infant mortality rate with and without vaccines(Sheet1).csv")

# Check the column names to make sure they match
head(data)

# Reshape data to long format
data_long <- data %>%
  pivot_longer(
    cols = c(`Vaccinated Mortality Rate`, `Unvaccinated Mortality Rate`),
    names_to = "Condition",
    values_to = "MortalityRate"
  )

# Plot both lines with linear regression fits
ggplot(data_long, aes(x = Year, y = MortalityRate, color = Condition)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(
    title = "Global Infant Mortality Rates Over Time",
    x = "Year",
    y = "Infant Mortality Rate (per 1,000 births)",
    color = "Condition"
  ) +
  scale_color_manual(values = c("Vaccinated Mortality Rate" = "purple", "Unvaccinated Mortality Rate" = "green"))
```

## Second Analysis:

# DISCUSSION

## First Analysis: Interpretation


## Second Analysis: Interpretation


# CONCLUSION


# REFERENCES



